%\VignetteEngine{knitr::knitr}
# nplr Vignette
### **AUTHORS**: _Frederic Commo (fredcommo@gmail.com) & Brian M. Bot_

```{r, include=FALSE, eval=TRUE}
require(nplr)
require(RCurl)
```

## Overview

This function computes a weighted n-parameters logistic regression, n from 2 to 5.  
Typical applications are drug-response or progression curve fitting.

The n-parameter logistic regression used by $nplr()$ is of the form:

$$y = B + \frac{(T - B)}{[1 + 10^{b.(xmid - x)}]^s}$$

Where $B$ and $T$ are the bottom and top asymptotes, respectively, $b$ and $xmid$ are the Hill slope and the x-coordinate at the inflexion point, respectively, and $s$ is an asymetric coefficient. This equation is sometimes refered to as a 5-parameter logistic regression, or the Richards equation.  
The $npars$ argument allows a user to run simplest models, while the default value $npars = "all"$ asks the function to test which model fits the best the data, with respect to a weighted Goodness-of-Fit estimator. See the $nplr$ documentation for more details.

In a drug-response (or progression) curve fitting context, typical needs are to invert the function in order to estimate the x-value, e.g. an IC50, given a y-value (the 0.5 survival rate). To do so, the implemented $predict()$ method takes 2 arguments: the model (an instance of the class nplr), and one (or a vector of) target(s).$predict()$ returns the corresponding x-values and their 95% confidence intervals, estimated by a Monte-Carlo simulation.

The $nplr()$ function has been optimized for fitting curves on y-values passed as proportions, from 0 to 1. If data are provided as original response values, e.g. optic density measurements, the $convertToProp()$ function may be helpful. For some curve fitting, the x-values may need to be log-transformed, e.g. drug concentrations. In such case, using $useLog=TRUE$ in $nplr()$ applies a $Log_10$ transformation on the x-values. Other arguments are described in the $nplr$ documentation.

A specific $plot()$ function has been implemented in order to visualize the results, using predefined plotting parameters. An easy way to draw simplest, or customized, plots is described in the examples below.

Several self-explanatory $get$ functions give an easy access to the results, model parameters and model performances stored in the $nplr()$ output.


```{r}
sessionInfo()
```

## Examples
```{r, simpleExample, fig.align='center', fig.width=9}
# A simple example: a drug-response fitting without replicates
url <- getURL("https://raw.github.com/fredcommo/IC50new/master/demo_files/demo1.tsv")
dat <- read.delim(text=url)
np1 <- nplr(dat$x, dat$y)
plot(np1)

# Getting the AUC and the x-estimates
getAUC(np1)
getEstimates(np1)

# Drawing a simple plot
op <- par(no.readonly=TRUE)
par(mfrow=c(1,2))
plot(np1)
plot(getX(np1), getY(np1))
lines(getXcurve(np1), getYcurve(np1))
par(op)
```
The $getAUC()$ method returns the area under the curve (AUC), approximated by the trapezoidal rule and the Simpson's rule.  
The $getEstimates()$ method returns the x-values, and their 95% intervals, for a predefined set of responses, from .1 to .9. As shown in the next example, the $predict()$ method is more flexible.  
Note that, such model cannot approximate x from y values outside the 2 asymptotes. In such case, the x approximation corresponding to the minimal (or maximal) possible y value will be returned.

The following example is a drug-response analysis, with replicated drug concentrations.
```{r, duplicates, fig.align='center', fig.width=9}
# Here a similar example, with replicated measurements
url <- getURL("https://raw.github.com/fredcommo/IC50new/master/demo_files/demo2.tsv")
dat <- read.delim(text=url)
np2 <- nplr(dat$x, dat$y)
plot(np2)
predict(np2, c(.25, .5, .75))
```

The next example shows the use of $nplr$ for fitting progression data.
```{r, progressCurve, fig.align='center', fig.width=9}
# An example of progression curve fitting
# where the x-values Log10 transformation is not required
# and the y-values are in some arbitrary scale.
url <- getURL("https://raw.github.com/fredcommo/IC50new/master/demo_files/demo4.tsv")
dat <- read.delim(text=url)

# convert the y-values to proportions
x <- dat$x
yp <- convertToProp(dat$y, 5, 110)
np3 <- nplr(x, yp, useLog=FALSE)
plot(np3, showTarget=FALSE, xlab="Time (hrs)", ylab="Progression")

# Getting the inflexion point coordinates
getInflexion(np3)

# Estimating the x_values, given a vector of responses
predict(np3, c(.25, .50, .75))
```
When a 5-p logistic regression is used, and because of the asymetric parameter, the curve is no longer symetrical around its inflexion point: this point can differs both from the mid-point and from the $y_{0.5}$ point.  
However, the inflexion point can be, in some situation, an important information to consider as it represents the coordinates at the maximum slope.  The $getInflexion$ gives access to its coordinates.

This last example shows the effect of the number of parameters on the curve fitting:
```{r, nparsEffect, fig.align='center', fig.width=9}
# On the same data: how the number of parameters can affect the fitting
url <- getURL("https://raw.github.com/fredcommo/IC50new/master/demo_files/demo4.tsv")
dat <- read.delim(text=url)

# convert the y-values to proportions
x <- dat$x
yp <- convertToProp(dat$y, 5, 110)
models <- lapply(2:5, function(p){
  tmp <- nplr(x, yp, useLog=FALSE, npars=p)
  list(x=getXcurve(tmp), y=getYcurve(tmp), infp=getInflexion(tmp), gof=getGoodness(tmp))
  })

plot(x, yp, col="grey", pch=19, cex=1.25)
for(i in 1:length(models)) {
  tmp <- models[[i]]
  lines(tmp$x, tmp$y, lwd=5, col=i)
  points(tmp$infp, pch=19, col=i, cex=2)
  legend(0, 1-i/10, legend=sprintf("%s-par: gof=%s", i+1, round(tmp$gof, 3)), lwd=2, col=i, bty="n")
  }
```


## References
Richards, F. J. (1959). A flexible growth function for empirical use. J Exp
Bot 10, 290–300.

Giraldo J, Vivas NM, Vila E, Badia A. Assessing the (a)symmetry of
concentration-effect curves: empirical versus mechanistic models. Pharmacol Ther.
2002 Jul;95(1):21-45.

Motulsky HJ, Brown RE. Detecting outliers when fitting data with nonlinear
regression - a new method based on robust nonlinear regression and the false
discovery rate. BMC Bioinformatics. 2006 Mar 9;7:123.
